---
title: "Project 3"
format: html
editor: visual
---

## Introduction and Data

The research question asks if the method of exoplanet discovery is correlated with the parameters present in the current exoplanet population. The method used to answer this question will mainly be k-means clustering. From the dataset, the majority of the variables were removed such as discovery date were removed as they were not relevant to the project. What was left was the name, discovery method, radius, and other exoplanet parameters.

Multivariate Analysis:

The visualization below shows a density graph of the orbital period comparing the radial velocity and transit. The graph shows that both have a peak around 25 days, suggesting that the exoplanet discovery methods favor a certain range of orbital periods. It should be noted that radial velocity has a larger range of days while transit's range is more focused.

```{r}
install.packages("ggplot2")
install.packages("tidyr")
install.packages("dplyr")
library(dplyr)
library(ggplot2)
library(tidyr)
filtered_data <- planetdata_cleaner %>%
  filter(discoverymethod %in% c("Transit", "Radial Velocity")) %>% # use these two because they encompass the majority of the data points
  select(pl_name, discoverymethod, pl_orbper, pl_massj, pl_orbsmax) %>% #filters the data so only these columns are present
  filter(!is.na(pl_orbper)) 
ggplot(filtered_data, aes(x = pl_orbper, fill = discoverymethod)) +
  geom_density(alpha = 0.5) +
  scale_x_log10() +   
  labs(title = "Orbital Period Distribution by Discovery Method",
       x = "Orbital Period (days, log scale)",
       y = "Density",
       fill = "Discovery Method") +
  theme_minimal() #plots a density map of orbital period vs. the filtered discovery methods
```

## Methodology

K-means clustering is the analysis that is currently being used as it serves to group together variables to find any patterns that might be present. In the code below, the variables representing orbital period, semi major axis, planetary mass, and planetary radius are being used to test k-means clustering.

```{r}
library(ggplot2)
library(dplyr)
library(datadictionary)
url <- "https://exoplanetarchive.ipac.caltech.edu/TAP/sync?query=select+*+from+pscomppars&format=csv"
planetdata <- read.csv(url) 
glimpse(planetdata) 
create_dictionary(planetdata)
planetdata_edit <- select(planetdata, "pl_name", "hostname", "discoverymethod", "pl_orbper", "pl_orbsmax", "pl_massj", "pl_eqt", "pl_orbeccen", "pl_rade") 
planetdata_cleaner <- na.omit(planetdata_edit)
planetdata_final <- subset(planetdata_cleaner, discoverymethod %in% c("Transit", "Radial Velocity", "Microlensing", "Imaging"))
planetdata_final$discoverymethod_label <- as.numeric(factor(planetdata_final$discoverymethod))
planetdata_final |> select(pl_orbper, pl_orbsmax) |> cor(use = "pairwise.complete.obs") |> round(3)
ggplot(planetdata_final, aes(pl_orbper, pl_massj, color = discoverymethod, shape = discoverymethod)) + geom_point(alpha = 0.25, size =1.5) + scale_x_log10() + scale_y_log10() + xlab("Orbital Period") + ylab("Planet Mass")
kmeanscols <- planetdata_final[,c("pl_orbper", "pl_massj")]
set.seed(123)
km.out <- kmeans(kmeanscols, centers = 4, nstart = 20)
km.out
planetdata_final$cluster_id <- factor(km.out$cluster)
ggplot(planetdata_final, aes(x = pl_orbper, y = pl_massj, color = cluster_id, shape = discoverymethod)) +
  geom_point(alpha = 0.5, size =3) + scale_x_log10()+
  xlab("Orbital Period") +
  ylab("Planet Mass")
km.out <- kmeans(kmeanscols, centers = 4, nstart = 20)
km.out
planetdata_final$cluster_id <- factor(km.out$cluster)
ggplot(planetdata_final, aes(x = pl_orbper, y = pl_rade, color = cluster_id, shape = discoverymethod)) +
  geom_point(alpha = 0.5, size =3) + scale_x_log10()+
  xlab("Orbital Period") +
  ylab("Planet Radius")
km.out <- kmeans(kmeanscols, centers = 4, nstart = 20)
km.out
planetdata_final$cluster_id <- factor(km.out$cluster)
ggplot(planetdata_final, aes(x = pl_orbsmax, y = pl_rade, color = cluster_id, shape = discoverymethod)) +
  geom_point(alpha = 0.5, size =3) + scale_x_log10()+
  xlab("Semi Major Axis") +
  ylab("Planet Radius")
km.out <- kmeans(kmeanscols, centers = 4, nstart = 20)
km.out
planetdata_final$cluster_id <- factor(km.out$cluster)
ggplot(planetdata_final, aes(x = pl_orbsmax, y = pl_massj, color = cluster_id, shape = discoverymethod)) +
  geom_point(alpha = 0.5, size =3) + scale_x_log10()+
  xlab("Semi Major Axis") +
  ylab("Planet Mass")
```

yYou can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

## Results

The five plots above highlight certain biases present in each method of exoplanet detection. The first plot shows a graph between planet mass and orbital period. No alterations are done to this plot except for applying the log scale to both axes'. The latter four plots all have gone through k-means clustering prior. All four share the pattern of the transit method favoring low period and low semi major axis planets. The transit method also favors low mass planets. Transiting planets have a much wider range in planet radius relative to the other detection methods. Imaging, microlensing, and radial velocity all have larger ranges in all categories except for planet radius.

These biases most likely exist due to the processes present in each exoplanet detection method. Transiting involves using the dip in a star's luminosity to detect an orbiting planet. As a result, it would be much easier to detect planets with a small orbital period as the dip in luminosity could be measured much more often. The methods used in radial velocity are independent of orbital period and semi major axis and instead focuses on how a planet shifts a star with its own gravitational pull. As a result, the planets it discovers have more variable parameters.
